---
title: "Data Pre-Processing"
format: html
editor: visual
execute: 
  output: true
  echo: true
  message: false
---

Here, basic steps to transform intensive longitudinal data into the format needed for the DMBM are explained. For illustrative purposes, we simulate a data set below.

```{r}
#| include: false
########## Download needed packages
library(dplyr)
library(tidyr)
library(lubridate)
#########
set.seed(1)

```

Here, we demonstrate how measurement-burst data can be prepared for the Dynamic Measurement-Burst Model (DMBM). Because the empirical data used in the manuscript cannot be shared publicly, we simulate a small example dataset.

The goal of this section is to illustrate **two key preprocessing steps**:

**1. Transform burst data from long format into a semi-wide format**\
**2. Construct and adjust the `TINTERVAL` variable required by Mplus**

These steps ensure that: - **within-burst dynamics** can be modeled in long format - **between-burst dynamics of process features** can be modeled in wide format - Mplus does not waste computation time interpreting large gaps between bursts.

# Simulate toy example

As the manuscript uses 3rd party data we are not permitted to share publicly, we will simulate example data below.

```{r}
# ============================================================
# 1) SIMULATE "LONG FORMAT" measurement-burst data (toy example)
#    - 6 participants
#    - 3 bursts (W1/W2/W3)
#    - 5 ESM prompts per burst
#    - 2 example variables: confidence, depression
#    - Each burst spans TWO days
# ============================================================

n_id    <- 6
n_waves <- 3
k_prompts <- 5

ids <- sprintf("P%02d", 1:n_id)

wave_start <- tibble(
  W = 1:3,
  wave_gap_days = c(0, 30, 60)  # 30-day gaps between bursts
)

df_long <- expand_grid(
  UUID = ids,
  W    = 1:3,
  t    = 1:k_prompts
) %>%
  left_join(wave_start, by = "W") %>%
  group_by(UUID, W) %>%
  mutate(
    base_time = ymd_hms("2020-01-01 08:00:00", tz = "Europe/Paris") +
      days(wave_gap_days[1]),
    
    # First 3 prompts on Day 1, remaining on Day 2
    day_offset = if_else(t <= ceiling(k_prompts / 2), 0, 1),
    
    # Within-day spacing every 3 hours
    hour_offset = 3 * ((t - 1) %% ceiling(k_prompts / 2)),
    
    Date_Time = base_time +
      days(day_offset) +
      hours(hour_offset) +
      minutes(sample(0:20, 1))  # small person-specific offset
  ) %>%
  ungroup() %>%
  mutate(
    confidence = round(rnorm(n(), mean = 2.5 + 0.3 * W, sd = 0.8), 2),
    depression = round(rnorm(n(), mean = 4.0 - 0.2 * W, sd = 0.9), 2)
  ) %>%
  select(UUID, W, t, Date_Time, confidence, depression)

print(df_long, n = 20)
```

**Structure of measurement-burst data**

The simulated data illustrate the typical structure of measurement-burst designs:

-   measurement occasions ( t ) nested within bursts ( W )
-   bursts nested within persons ( UUID )

Importantly, bursts are separated by **large time gaps** (e.g., 30 days), while observations **within bursts are closely spaced** (e.g., every 3 hours).

This distinction is crucial when constructing the `TINTERVAL` variable.

# Data Transformation

To prepare the data for the DMBM, we first reshape the dataset into a **semi-wide format**.

In this format:

-   intensive longitudinal observations remain in **long format**

-   burst-level variables appear as **separate columns per burst**

This allows the model to simultaneously estimate:

-   within-burst dynamics

-   between-burst dynamics of process features.

**Step: Convert bursts to columns**

Each burst receives its own set of variables (e.g., `confidence_W1`, `confidence_W2`, `confidence_W3`).

```{r}
#####################################################################################################################
df_long <- df_long %>%
  group_by(UUID, W) %>%
  arrange(Date_Time, .by_group = TRUE) %>%
  mutate(t = row_number()) %>%   # 1..k per person×wave
  ungroup()

# transform to wide format:

df_wide <- df_long %>%
  # choose the variables that should become wave-specific columns
  pivot_wider(
    id_cols = c(UUID, t, Date_Time),
    names_from = W,
    values_from = c(confidence, depression),
    names_glue = "{.value}_W{W}"
  )

print(df_wide, n = 20)

#####################################################################################################################

```

# Adding `TINTERVAL` to the data

Dynamic models in Mplus (DSEM) require a continuous time variable that represents elapsed time since the beginning of the study for each participant. This variable is then referenced in the `TINTERVAL` option in Mplus.

Instead of counting measurement occasions, this variable encodes **time since the first observation for each participant**.

For example:

-   first observation → `0`\

    second observation 3 hours later → `3`\

    third observation 6 hours later → `6`

This allows Mplus to correctly interpret temporal precedence and to correctly account for unequal spacing between observations.

------------------------------------------------------------------------

## Measurement-burst complication

Measurement-burst data introduce a practical complication:

The time gap between bursts can be extremely large.\
For example, a 30-day gap corresponds to **720 hours**.

If left unchanged, Mplus interprets these large gaps literally and inserts many missing time points internally, which can dramatically slow estimation.

------------------------------------------------------------------------

## Key idea

**We only model dynamics within bursts.**

Therefore:

-   long gaps **between bursts do not carry meaningful information** -

-   compressing them **does not affect estimation**

We therefore:

-   keep the temporal ordering within bursts intact

-   replace large inter-burst gaps with a small fixed interval (e.g., 24 hours).

This preserves the temporal structure needed for estimation while avoiding unnecessary computational burden.

This can be done in 3 steps:

### Step 1 — Create time interval variable

We first compute elapsed time since each participant’s first observation.

```{r}
#####################################################################################################################

df_time <- df_long %>%
  group_by(UUID) %>%
  arrange(Date_Time, .by_group = TRUE) %>%
  mutate(
    seconds_since_start_pp = as.numeric(Date_Time - first(Date_Time), units = "secs"),
    hours_since_start_pp   = seconds_since_start_pp / 3600
  ) %>%
  ungroup()

df_time %>% select(UUID, W, t, Date_Time, hours_since_start_pp) %>% head(10)

```

### Step 2 — Identify burst boundaries per participant

We determine the first and last observation of each burst.

```{r}
#adjust gaps between waves:#
# 1) Get per-person wave boundaries (first/last Date_Time per wave)
pp_bounds <- df_long %>%
  group_by(UUID, W) %>%
  summarise(
    W_first = min(Date_Time),
    W_last  = max(Date_Time),
    .groups = "drop"
  ) %>%
  pivot_wider(
    id_cols = UUID,
    names_from = W,
    values_from = c(W_first, W_last),
    names_glue = "W{W}_{.value}"
  )

pp_bounds

```

### Step 3 — Compress gaps between bursts

Large inter-burst gaps are replaced with a fixed interval (24 hours here).

```{r}
# 2) Join bounds back to each row, compute adjusted seconds
df_adj <- df_time %>%
  left_join(pp_bounds, by = "UUID") %>%
  mutate(
    seconds_since_start_adjusted = case_when(
      W == 1 ~ seconds_since_start_pp,
      
      # Wave 2 starts 24h after Wave 1 ended (for that participant)
      W == 2 ~ {
        gap_W1_W2 <- as.numeric(W2_W_first - W1_W_last, units = "secs")
        seconds_since_start_pp - gap_W1_W2 + 86400
      },
      
      # Wave 3 starts 24h after Wave 2 ended
      W == 3 ~ {
        gap_W1_W2 <- as.numeric(W2_W_first - W1_W_last, units = "secs")
        gap_W2_W3 <- as.numeric(W3_W_first - W2_W_last, units = "secs")
        
        # start from pp time, compress W2->W3 gap to 24h, and keep W2 positioned 24h after W1
        seconds_since_start_pp - gap_W2_W3 + 86400 - gap_W1_W2 + 86400
      }
    ),
    tinterval_h = round(seconds_since_start_adjusted / 3600, 2) # transform to hours
  )

df_adj %>% select(UUID,Date_Time, W, t, hours_since_start_pp, tinterval_h) %>% head(12)
```

### Final dataset

The resulting dataset:

-   preserves within-burst timing

-   removes large inter-burst gaps

-   is ready for DSEM estimation in Mplus.

```{r}
adjusted_df <- df_adj %>%
  # example: keep ID, wave, adjusted time, and wave-specific outcomes
  pivot_wider(
    id_cols = c(UUID,Date_Time, t, tinterval_h),
    names_from = W,
    values_from = c(confidence, depression),
    names_glue = "{.value}_W{W}"
  ) %>%
  # drop rows where ALL outcome columns are NA
  filter(if_any(starts_with("confidence_") | starts_with("depression_"), ~ !is.na(.x)))

print(adjusted_df,n=20)
```